MS ANXIETY

### 2025-01-20 scripts/jama_scripts/MSAnxiety_manuscript_JAMA_revision_20250116.Rmd
clean

 #print summary stats
 gam_summary_stats(volume_of_mimosa_lesions_gam)
```

#### Anxiety dose
  - anxiety dose correlates with disease burden (volume of lesions)
```{r lesion_volumes_anxiety_dose}
df_demo_and_fascicles_no_unclass_anxiety_dose <- df_demo_and_fascicles %>%
  filter(oanxietydx_OR_meds_AND_Anxietydx.Meds.And.Dep.True.Healthy.Collapsed != "unclassified") %>%
  filter (anxiety_dose != -1) #n=372

@@

   #print summary stats
 gam_summary_stats(anxiety_dose_volume_of_mimosa_lesions_by_sex_gam)

@@

  corr_LV_to_UF <- cor.test(uncinate_by_dx_df$volume_of_mimosa_lesions,uncinate_by_dx_df$mean_UF_vol) #corr 0.69; 
  print(corr_LV_to_UF)

  lesion_vol_UF_lm <- lm(uncinate_by_dx_df$volume_of_mimosa_lesions~ uncinate_by_dx_df$mean_UF_vol) #P <2.2*10^-16, they are colinear
  summary(lesion_vol_UF_lm) #T=10.585

@@

  ## Try scaling mean volume
  disease_mean_scaled_UF_vol_gam <- gam(mean_UF_vol_scaled_total_vol_lesions ~ Diagnosis + osex+s(PAT_AGE_AT_EXAM, k = 4, fx = F), data = uncinate_by_dx_df)
  summary(disease_mean_scaled_UF_vol_gam) #- N.S.

  #print summary stats; (T = -0.556, P = 0.579, Cohen's f2 =  0.007, Effect size = minimal)"
  gam_summary_stats(disease_mean_scaled_UF_vol_gam)

  #How about predicting Diagnosis? - nope
  diagnosis_prediction <- glm(Diagnosis ~ mean_UF_prop + Diagnosis + osex+PAT_AGE_AT_EXAM + volume_of_mimosa_lesions, data = uncinate_by_dx_df, family = "binomial")
  summary(diagnosis_prediction)
  confint(diagnosis_prediction)

  #how about predicting diagnosis with scaled UF? N.S.
  diagnosis_prediction_scaled <- glm(Diagnosis ~ mean_UF_vol_scaled_total_vol_lesions + Diagnosis + osex+PAT_AGE_AT_EXAM, data = uncinate_by_dx_df, family = "binomial")
  summary(diagnosis_prediction_scaled)
  confint(diagnosis_prediction_scaled)

@@

    #bilat anxiety dose + sex + age: anxiety_dose    0.04989    0.02397   2.082   0.0381 * original result

  #How about predicting Diagnosis? volume_of_mimosa_lesions  1.518e-05  7.147e-06   2.124   0.0357 *
  diagnosis_prediction <- lm(anxiety_dose ~ mean_UF_prop + osex+PAT_AGE_AT_EXAM + volume_of_mimosa_lesions, data = uncinate_by_dx_df)
  summary(diagnosis_prediction)
  confint(diagnosis_prediction)

### modeling volume of mimosa lesions like this doesn't make sense
 #anxietyGrouping_mean_UF_by_sex_w_LV_gam_prop <- gam(mean_UF_prop ~ anxiety_dose + osex + s(PAT_AGE_AT_EXAM, k = 4, fx = F) + volume_of_mimosa_lesions, data=df_demo_and_fascicles_no_unclass_anxiety_dose)
 # summary(anxietyGrouping_mean_UF_by_sex_w_LV_gam_prop)

  #print summary stats
  #gam_summary_stats(anxietyGrouping_mean_UF_by_sex_w_LV_gam_prop)

@@

  total_vol_UF_lm <- lm(uncinate_by_dx_df$total_volume~ uncinate_by_dx_df$mean_UF_vol) #P <2.2*10^-16, they are colinear
  summary(total_vol_UF_lm)

@@

  total_vol_UF_lm <- lm(df_demo_and_fascicles_no_unclass_anxiety_dose$total_volume~ df_demo_and_fascicles_no_unclass_anxiety_dose$mean_UF_vol) #P <2.2*10^-16, they are colinear
  summary(total_vol_UF_lm)

@@

```
##### JAMA Analysis: Residuals vs UF
```{r dep_anxiety_UF}

## Definte data frams
################
## depression ##
################
df_dep <- df_demo_and_fascicles %>%
  filter(depGroupVar != 0) %>%
  mutate(depDiagnosis = as.factor(ifelse(depGroupVar ==2, "MS+Depression", "MS+noDepression")))

####################### 
## anxiety diagnosis ##
#######################
df_anxiety_dx <- df_demo_and_fascicles %>%
  filter(oanxietydx_OR_meds_AND_Anxietydx.Meds.And.Dep.True.Healthy.Collapsed != "unclassified") %>%
  filter(oanxietydx_OR_meds_AND_Anxietydx.Meds.And.Dep.True.Healthy.Collapsed != "Has.Anxietydx.OR.Antianxiety.Meds") %>%
  mutate(subject = as.factor(EMPI)) %>%
  mutate(anxiety_Diagnosis = oanxietydx_OR_meds_AND_Anxietydx.Meds.And.Dep.True.Healthy.Collapsed) %>%
  mutate(anxiety_Diagnosis = ifelse(oanxietydx_OR_meds_AND_Anxietydx.Meds.And.Dep.True.Healthy.Collapsed=="true_healthy", "MS-Anxiety", "MS+Anxiety")) %>%
  mutate(anxiety_Diagnosis = as.factor(anxiety_Diagnosis)) %>%
  ungroup()

##################
## anxiety dose ##
################## 
df_anxiety_dose <-  df_demo_and_fascicles %>%
  filter(oanxietydx_OR_meds_AND_Anxietydx.Meds.And.Dep.True.Healthy.Collapsed != "unclassified") %>%
  filter(anxiety_dose != -1) #n=372

### stats
# Step 1: Fit the three models

# Model 1: depDiagnosis with df_dep
model_depDiagnosis <- gam(mean_UF_prop ~ depDiagnosis + s(PAT_AGE_AT_EXAM, k=4, fx=F) + osex, data = df_dep)

# Model 2: anxiety_Diagnosis with df_anxiety_dx
model_anxiety_Diagnosis <- gam(mean_UF_prop ~ anxiety_Diagnosis + s(PAT_AGE_AT_EXAM, k=4, fx=F) + osex, data = df_anxiety_dx)

# Model 3: anxiety_dose with df_anxiety_dose
model_anxiety_dose <- gam(mean_UF_prop ~ anxiety_dose + s(PAT_AGE_AT_EXAM, k=4, fx=F) + osex, data = df_anxiety_dose)

# Step 2: Extract residuals from each model (mean_UF_prop adjusted for age and sex)
residuals_depDiagnosis <- residuals(model_depDiagnosis)
residuals_anxiety_Diagnosis <- residuals(model_anxiety_Diagnosis)
residuals_anxiety_dose <- residuals(model_anxiety_dose)

# Step 3: Calculate the correlation between mean_UF_prop and the predictors (residuals)
cor_depDiagnosis <- cor(df_dep$mean_UF_prop, residuals_depDiagnosis)
cor_anxiety_Diagnosis <- cor(df_anxiety_dx$mean_UF_prop, residuals_anxiety_Diagnosis)
cor_anxiety_dose <- cor(df_anxiety_dose$mean_UF_prop, residuals_anxiety_dose)

# Print the correlation results
cat("Correlation between mean_UF_prop and depDiagnosis: ", cor_depDiagnosis, "\n")
cat("Correlation between mean_UF_prop and anxiety_Diagnosis: ", cor_anxiety_Diagnosis, "\n")
cat("Correlation between mean_UF_prop and anxiety_dose: ", cor_anxiety_dose, "\n")

# Step 4: Perform Steiger's test for comparing the correlations
# Create a function for Steiger's test (comparison of two correlated correlations)
steiger_test <- function(r12, r13, n) {
  # r12: correlation between variable1 and variable2
  # r13: correlation between variable1 and variable3
  # n: sample size

  # Calculate the Z score for comparison
  z <- (r12 - r13) / sqrt((1 - r12^2) / (n - 3) + (1 - r13^2) / (n - 3))
  p_value <- 2 * (1 - pnorm(abs(z)))  # Two-tailed p-value

  return(list(Z_score = z, p_value = p_value))
}

# Sample size (number of subjects)
n = length(Reduce(intersect, list(df_dep$EMPI, df_anxiety_dx$EMPI, df_anxiety_dose$EMPI)))

# Step 5: Test for differences between correlations
# Test if the correlation between anxiety_Diagnosis and mean_UF_prop is stronger than depression
steiger_test_anxiety_vs_dep <- steiger_test(cor_anxiety_Diagnosis, cor_depDiagnosis, n)

# Test if the correlation between anxiety_dose and mean_UF_prop is stronger than depression
steiger_test_anxiety_dose_vs_dep <- steiger_test(cor_anxiety_dose, cor_depDiagnosis, n)

# Output the results of Steiger's test; Z score:  -0.7806441 ; p-value:  0.4350118 
cat("Steiger's Test: Anxiety Diagnosis vs Depression Diagnosis\n")
cat("Z score: ", steiger_test_anxiety_vs_dep$Z_score, "\n")
cat("p-value: ", steiger_test_anxiety_vs_dep$p_value, "\n")

#Z score:  0.07445476 ; p-value:  0.9406485 
cat("Steiger's Test: Anxiety Dose vs Depression Diagnosis\n")
cat("Z score: ", steiger_test_anxiety_dose_vs_dep$Z_score, "\n")
cat("p-value: ", steiger_test_anxiety_dose_vs_dep$p_value, "\n")
```

```
##### JAMA ANALYSIS - compare correlations w/wout uncinate
```{r other_injured_fascicles}
df_demo_and_fascicles_no_unclass_anxiety_dose <- df_demo_and_fascicles %>%
  filter(oanxietydx_OR_meds_AND_Anxietydx.Meds.And.Dep.True.Healthy.Collapsed != "unclassified") %>%
  filter(anxiety_dose != -1)  %>% #n=372 
  mutate(total_lesion_burden_no_UF = sum_fascicle_vol_lost - (UF_L_vol + UF_R_vol))

#almost identical because uncinate is small
#t=2.05, p = 0.04083
cor_lesion_burden_by_anxiety = cor.test(df_demo_and_fascicles_no_unclass_anxiety_dose$anxiety_dose, df_demo_and_fascicles_no_unclass_anxiety_dose$sum_fascicle_vol_lost)
print(cor_lesion_burden_by_anxiety)

#t=2.05, p = 0.04075
cor_lesion_burden_no_UF_by_anxiety = cor.test(df_demo_and_fascicles_no_unclass_anxiety_dose$anxiety_dose, df_demo_and_fascicles_no_unclass_anxiety_dose$total_lesion_burden_no_UF)
print(cor_lesion_burden_no_UF_by_anxiety)

@@

##### Overall disease in brain
```{r}
#n = 372

>>>>

 #print summary stats; (T = 3.04, P = 0.003, Cohen's f2 =  0.342, Effect size = medium)
 gam_summary_stats(volume_of_mimosa_lesions_gam)
```

#### Anxiety dose- anxiety dose correlates with disease burden (volume of lesions)
```{r lesion_volumes_anxiety_dose}
df_demo_and_fascicles_no_unclass_anxiety_dose <- df_demo_and_fascicles %>%
  filter(oanxietydx_OR_meds_AND_Anxietydx.Meds.And.Dep.True.Healthy.Collapsed != "unclassified") %>%
  filter (anxiety_dose != -1) #n=372

@@

   #print summary stats; (T = 3.084, P = 0.0022, Cohen's f2 =  0.12, Effect size = small)
 gam_summary_stats(anxiety_dose_volume_of_mimosa_lesions_by_sex_gam)

@@

  corr_LV_to_UF <- cor.test(uncinate_by_dx_df$volume_of_mimosa_lesions,uncinate_by_dx_df$mean_UF_vol) #corr 0.69; p-value < 2.2e-16
  print(corr_LV_to_UF)

@@

  ## Try scaling mean volume
 # disease_mean_scaled_UF_vol_gam <- gam(mean_UF_vol_scaled_total_vol_lesions ~ Diagnosis + osex+s(PAT_AGE_AT_EXAM, k = 4, fx = F), data = uncinate_by_dx_df)
  #summary(disease_mean_scaled_UF_vol_gam) #- N.S.

  #print summary stats; (T = -0.556, P = 0.579, Cohen's f2 =  0.007, Effect size = minimal)"
  #gam_summary_stats(disease_mean_scaled_UF_vol_gam)

@@

corr_LV_to_UF_anxiety_dose <- cor.test(df_demo_and_fascicles_no_unclass_anxiety_dose$volume_of_mimosa_lesions,df_demo_and_fascicles_no_unclass_anxiety_dose$mean_UF_vol) #
  print(corr_LV_to_UF_anxiety_dose)

@@

@@

@@

@@

##### Overall disease in brain
```{r overall_disease_bruden}
#n = 372
